---
title: "STAT321_FinalReport"
output:
  html_document:
    df_print: paged
date: "2023-06-08"
---

# FINAL Report: Investigating The Relationship Between Land Use Diversity and Pedestrian Activity in Seattle
### Authors: Finch Brown and JP Lopez

## Introduction and Research Design
The primary research question guiding this investigation is: Does land use diversity influence the level of pedestrian activity in urban spaces? This study's focus stems from the conjecture that diversity in land use could be a significant factor influencing the movement of pedestrians. From a theoretical perspective, diverse land uses offer more opportunities for activity and thus draw more pedestrian traffic. Therefore, our hypothesis for this study is as follows:

H0: Land use diversity has no significant impact on pedestrian activity. (β_diversity = 0)
H1: Land use diversity positively influences pedestrian activity. (β_diversity > 0)

The importance of this research lies in its potential implications for urban development and planning, ultimately encouraging active and sustainable urban environments.

The research design for this investigation utilizes observational data from real-world observations in Seattle. We have categorized each location according to the classifications used by the Gehl Institute's Public Life Data Protocol, distinguishing them into single-use or mixed-use areas. Our measurement of pedestrian activity is based on total counts of moving pedestrians recorded in each entry of the dataset.

If the data analysis reveals a significant positive relationship between land use diversity and pedestrian activity, it will provide empirical support for our hypothesis, underscoring the contribution of diverse land use to increased pedestrian movement.

## Data Cleaning
### A data section describing the source of the data, measurement, and uses plots to summarize the dependent variable
### Exploratory Data Analysis: analysis: Let the data speak by constructing relevant and polished plots or visuals that reveal important patterns and relationships related to the research question and hypotheses. Provide a narrative explanation of these patterns and their potential influence on the model specification.


```{r}

## Prerequisite
### Load Libraries
library(tidyverse)
library(dplyr)
library(ggplot2)

### loading datasets
people_moving_df <- read.csv("Public_Life.csv")
View(people_moving_df)

locations_df <- read.csv("locations.csv")
View(locations_df)

```


The datasets used in this report were sourced from the Public Life Data repository by Seattle's Department of Transportation (link: https://data.seattle.gov/Transportation/Public-Life-Data-People-Moving/7rx6-5pgd). The primary dataset, 'Public Life Data: People Moving,' comprises pedestrian counts from different Seattle locations. The second dataset, 'Locations,' contains various land use types (e.g., Residential, Mixed, Commercial, etc.).

To illustrate the dependent variable, i.e., pedestrian movement, a boxplot is presented, demonstrating pedestrian counts in single-use and mixed-use spaces.

```{r}
### isolate relevant columns
unique(locations_df$location_character)
small_loc_df <- data.frame(locations_df$location_id, locations_df$location_line_typology_vehicular)
small_move_df <- data.frame(people_moving_df$unique_moving_id, people_moving_df$moving_row_total, people_moving_df$moving_time_start, people_moving_df$moving_time_end, people_moving_df$location_id)

### rename columns
colnames(small_loc_df) <- c("location", "typology")
colnames(small_move_df) <- c("id", "total_count", "start", "end", "location")

### join the dataframes
new_df <- left_join(small_loc_df, small_move_df, by = join_by(location), multiple = "all")

### creating column to categorize neighborhood types
unique(new_df$typology)

### classify each location as either single-use or mixed-use
new_df$classification <- ifelse(new_df$typology %in% c("Neighborhood Corridor", "Urban Village Neighborhood Access", "Urban Village Neighborhood", "Downtown Neighborhood", "Downtown", "Downtown Neighborhood Access", "Urban Center Connector", "Urban Village Main"), 1, 0)

### add a column for count per unit of time
new_df$start <- as.POSIXlt(new_df$start, format="%d/%m/%Y %I:%M:%S %p", tz = "UTC", na.rm = T)
new_df$end <- as.POSIXlt(new_df$end, format="%d/%m/%Y %I:%M:%S %p", tz = "UTC", na.rm = T)
new_df$duration <- as.numeric((new_df$end - new_df$start) / 60)
new_df$count_per_minute <- new_df$total_count / new_df$duration

### add a column to label land use
new_df$land_use <- ifelse(new_df$classification == 1, "Mixed Use", "Single Use")

### remove NA rows
new_df_clean <- na.omit(new_df)

### Install and load the 'moments' package
library(moments)

## Exploratory Data Analysis
### Descriptive Statistics:
### Compute variance, standard deviation, skewness, and kurtosis
variance <- var(new_df_clean$total_count)
standard_deviation <- sd(new_df_clean$total_count)
skewness <- skewness(new_df_clean$total_count)
kurtosis <- kurtosis(new_df_clean$total_count)

### Print the computed statistics
print(paste('Variance: ', variance))
print(paste('Standard deviation: ', standard_deviation))
print(paste('Skewness: ', skewness))
print(paste('Kurtosis: ', kurtosis))

### Correlation matrix among the quantitative variables
correlation_matrix <- cor(new_df_clean[,sapply(new_df_clean, is.numeric)])
print(correlation_matrix)

### Boxplot
ggplot(new_df_clean, aes(x = land_use, y = total_count, fill = land_use)) +
  geom_boxplot() +
  xlab("Land Use Diversity") +
  ylab ("Number of People Moving") +
  theme(legend.position = "none")

### Violin plot with different colors for single-use and mixed-use
ggplot(new_df_clean, aes(x = land_use, y = total_count, fill = land_use)) +
  geom_violin() +
  geom_boxplot(width=0.1, fill="white") +
  xlab("Land Use Diversity") +
  ylab ("Number of People Moving") +
  theme(legend.position = "none")

```


## Section Three: Results
### A results section containing a scatterplot of the main relationship of interest and output for the main regression

## Statistical Analysis
### Bivariate regression: total_count ~ classification


```{r}
model1 <- lm(total_count ~ classification, data = new_df_clean)
summary(model1)  

# Add controls progressively and present the full specification?
# model2 <- lm(total_count ~ classification + control1, data = new_df_clean)
# summary(model2)

# model3 <- lm(total_count ~ classification + control1 + control2, data = new_df_clean)
# summary(model3)

# Final model with all controls
# model4 <- lm(total_count ~ classification + control1 + control2 + control3, data = new_df_clean)
# summary(model4)
```
To illustrate the main relationship of interest, a scatterplot is provided below. The scatterplot showcases the distribution of pedestrian counts relative to land use diversity.

```{r}
# remove rows with NAs
new_df_clean <- na.omit(new_df)

# generate the scatterplot
ggplot(new_df_clean, aes(x = land_use, y = total_count, color = land_use)) +
  geom_jitter(width = 0.3) +
  xlab("Land Use") +
  ylab("Pedestrians Through the Space")
```
The principal regression analysis was performed with the 'total_count' as the outcome variable and 'classification' (land use diversity) as the explanatory variable. The regression output is as follows:

```{r}

model <- lm(total_count ~ classification, data = new_df)
summary(model)

```

The model indicates a significant positive relationship between land use diversity and pedestrian activity. This relationship suggests that a higher level of land use diversity corresponds to an increase in pedestrian movement.

## Conclusion
### A brief conclusion summarizing the results, assesses the extent to which you found support for your hypothesis, and describes limitations of your analysis and threats to inference
The analysis provides evidence supporting our hypothesis that land use diversity positively influences pedestrian activity in Seattle. This finding has significant implications for urban development strategies and the promotion of pedestrian-friendly environments.

However, these results should be interpreted with some limitations in mind. The study is observational and, as such, cannot confirm causality. Additionally, the model explains a relatively small portion of the variance in pedestrian activity, indicating that other factors beyond land use diversity likely contribute significantly to pedestrian movement.

Moving forward, research may consider incorporating more variables into the analysis, such as population density, available amenities, and public transportation facilities, to gain a more comprehensive understanding of pedestrian activity determinants.

Despite the limitations, this analysis provides valuable insights into urban design and planning, hinting that land use diversity can foster more vibrant, pedestrian-friendly environments. The results of this study provide a stepping stone for future research exploring urban environments and pedestrian activity. 
